Fusing Filters with Integer Linear Programming

### 2014-05-14 2014betterfusionforfilters/section/03-SizeInference.tex
transducer changes
varied fixes
remove those damn figure[H] because they were reordering things strangely

Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
\begin{alltt}
filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
  Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}

However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process. In our situation, we need automatic type generalization, but for a first order language only.

@@

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}

The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

@@

To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
\begin{tabbing}
MMMM \= M \= MMMMMMM \= M \= \kill
$parents$ \> $::$ \> $name \to name \to \{name \times name\}$ \\
$parents(a,b)$ \\
        \> $|$ \> $\tau(a) == \tau(b)$ \> $=$    \> $\{(a, b)\}$ \\
        \> $|$ \> otherwise            \> $=$    \> $\{ parents(a', b) ~|~ a' \in trans(a) \} $      \\
        \>     \>                      \> $\cup$ \> $\{ parents(a, b') ~|~ b' \in trans(b) \} $  \\
\end{tabbing}

\textbf{Lemma: sole transducers}.
For some bindings $bs$, each name $n$ will have at most one transducer.
\[
valid(\constr{bs}) \implies \forall n.\ |trans(bs, n)| \le 1
\]
% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
% The other cases are trivial.

\textbf{Lemma: transducers change types}.
For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
\[
valid(\constr{bs}) \wedge \{m\} = trans(bs, n) \implies \tau(n) \not= \tau(m)
\]

\textbf{Lemma: sole parents}.
For some bindings $bs$, each pair of names $a$ and $b$ will have at most one set of parents $parents(a,b)$.
\[
valid(\constr{bs}) \implies \forall a,b.\ |parents(bs, n)| \le 1
\]
% \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.

>>>>

Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
\begin{alltt}
 filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
   Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}
However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process. In our situation, we need automatic type generalization, but for a first order language only.

@@

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}
The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

@@

To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
\begin{tabbing}
MMMM \= M \= M \= M \= \kill
$parents$ \> $::$ \> $binds \to name \to name \to \{name \times name\}$ \\
$parents(bs, a, b)$ \\
        \> $|$ \> $\iiter_{\Gamma,C}(bs(a)) == \iiter_{\Gamma,C}(bs(b))$ \\
        \>     \>                      \> $=$ \> $\{(a, b)\}$ \\
        \> $|$ \> otherwise            \\
        \>     \>                      \> $=$    \> $\{ parents(bs, a', b) ~|~ a' \in trans(bs, a) \} $      \\
        \>     \>                      \> $\cup$ \> $\{ parents(bs, a, b') ~|~ b' \in trans(bs, b) \} $  \\
\end{tabbing}

\textbf{Lemma: sole transducers}.
For some bindings $bs$, each name $n$ will have at most one transducer.
\[
\forall bs, n.\ |trans(bs, n)| \le 1
\]
% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
% The other cases are trivial.

% this is only useful for proving sole parents
% \textbf{Lemma: transducers change types}.
% For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
% \[
% \forall bs, n, m.\ trans(bs, n) = \{m\} \implies \tau(n) \not= \tau(m)
% \]

\textbf{Lemma: sole parents}.
For some bindings $bs$, each pair of names $a$ and $b$ will have at most one set of parents $parents(a,b)$.
\[
\forall bs, a, b.\ |parents(bs, a, b)| \le 1
\]
% \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
remove those damn figure[H] because they were reordering things strangely

\begin{figure}[H]
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A2) (B1)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}
\end{figure}
For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}. The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@). For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@. For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

@@

\begin{figure}[H]
\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
\end{tabbing}
\end{figure}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{figure}[H]
\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
\\
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
    \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, from, \fusible)$            \\
\end{tabbing}
\end{figure}

@@

As an example of a cyclic clustering, consider:
\begin{figure}[H]
\begin{code}
cycle us
 = let xs  = map (+1) us          (C1)
       sum = fold xs              (C2)
       zs  = map (+sum) xs        (C1)
\end{code}
\end{figure}
There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from $i$ to $j$)            \\
           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from $i$ to $j$)            \\
           \> \ldots
\end{tabbing}
\end{figure}
As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.

@@

\paragraph{Fusion-preventing edges}
\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (fusion-preventing edges from $i$ to $j$)      \\
           \> \ldots
\end{tabbing}
\end{figure}
As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur. This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\}$) \\
           \> \ldots
\end{tabbing}
\end{figure}
If two nodes have different iteration sizes, they may not necessarily be fused.
If either of the nodes have no iteration size --- in which case, they are @external@ computations --- they may not be fused at all.
However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.

@@

As a degenerate example, consider fusing an operation on filtered data with its generating filter:
\begin{figure}[H]
\begin{code}
 sum1 = fold (+) 0  us
 gts  = filter (>0) us
 sum2 = fold (+) 0  gts
\end{code}
\end{figure}
Here, $parents(gts,sum2) = \{gts, gts\}$.
This generates spurious, but still valid, constraints that $gts$ must be fused with $gts$ and $gts$ must be fused with $sum2$, in order for $gts$ and $sum2$ to be fused together.
While these constraints are unnecessary in this case, they are harmless.

@@

\paragraph{Array contraction}
\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
\end{figure}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it.
Thus, $c_i=0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
           \> \> \> (memory traffic and loop overhead)         \\
           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
           \> \> \> (removing intermediate arrays)         \\
Subject to \> \ldots                                \\
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}
\end{figure}

@@

\begin{figure}[H]
\begin{tabbing}
MMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$split(a,b)$ \\
    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$\\
\end{tabbing}
\end{figure}

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \\
           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \> (if $split(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
\\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \> (if $split(i,j) \wedge (i,j,\fusible) \in E$)     \\
\\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \> (if $split(i,j) \wedge (i,j,\fusionpreventing) \in E$)    \\
\\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (if $split(i,j)$) \\
           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
           \> (if $\neg(split(i,j))$) \\
\\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\} \wedge$ \\

           \> \> $ split(i,j) \wedge split(i',j) \wedge split(i,j') \wedge split(i',j') $) \\
\end{tabbing}
\end{figure}

>>>>

\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A2) (B1)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}
For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}. The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@). For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@. For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

@@

\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
\end{tabbing}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
\\
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
    \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, from, \fusible)$            \\
\end{tabbing}

@@

As an example of a cyclic clustering, consider:
\begin{code}
cycle us
 = let xs  = map (+1) us          (C1)
       sum = fold xs              (C2)
       zs  = map (+sum) xs        (C1)
\end{code}
There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from $i$ to $j$)            \\
           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from $i$ to $j$)            \\
           \> \ldots
\end{tabbing}
As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.

@@

\paragraph{Fusion-preventing edges}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (fusion-preventing edges from $i$ to $j$)      \\
           \> \ldots
\end{tabbing}
As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur. This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\}$) \\
           \> \ldots
\end{tabbing}
If two nodes have different iteration sizes, they may not necessarily be fused.
If either of the nodes have no iteration size --- in which case, they are @external@ computations --- they may not be fused at all.
However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.

@@

As a degenerate example, consider fusing an operation on filtered data with its generating filter:
\begin{code}
 sum1 = fold (+) 0  us
 gts  = filter (>0) us
 sum2 = fold (+) 0  gts
\end{code}
Here, $parents(gts,sum2) = \{gts, gts\}$.
This generates spurious, but still valid, constraints that $gts$ must be fused with $gts$ and $gts$ must be fused with $sum2$, in order for $gts$ and $sum2$ to be fused together.
While these constraints are unnecessary in this case, they are harmless.

@@

\paragraph{Array contraction}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it.
Thus, $c_i=0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.

@@

\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
           \> \> \> (memory traffic and loop overhead)         \\
           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
           \> \> \> (removing intermediate arrays)         \\
Subject to \> \ldots                                \\
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}

@@

\begin{tabbing}
MMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$split(a,b)$ \\
    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$\\
\end{tabbing}

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \\
           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \> (if $split(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
\\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \> (if $split(i,j) \wedge (i,j,\fusible) \in E$)     \\
\\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \> (if $split(i,j) \wedge (i,j,\fusionpreventing) \in E$)    \\
\\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (if $split(i,j)$) \\
           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
           \> (if $\neg(split(i,j))$) \\
\\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\} \wedge$ \\

           \> \> $ split(i,j) \wedge split(i',j) \wedge split(i,j') \wedge split(i',j') $) \\
\end{tabbing}
